mathml: Translate R expressions to MathML and LaTeX/MathJax

### 2022-12-11 Draft3.Rmd
table

---
title: "mathml: Translate R expressions to MathML and LaTeX/MathJax"
date: "2022-12-09"
abstract: >
  This R\ package translates R\ objects to suitable entities in MathML or LaTeX,
  thereby enabling pretty mathematical representation of R\ functions or
  algorithms in data analysis, scientific reports and interactive web content.
  R\ code text descriptions and mathematical content already coexist,
  side-by-side, in the RMarkdown document rendering language. The present
  package allows to use the same R\ objects for both data analysis and
  typesetting in documents or web content. By tightening the link between the
  statistical analysis and its verbal or symbolic representation, this new
  feature is a futher step towards reproducible science. User-defined hooks
  allow to extend the package by mapping new variable names and functions to
  new MathML and LaTeX entities. We provide a few working examples that
  illustrate the use of the package in a little scientific report and in an
  interactive web page.
draft: true
author:  
  # see ?rjournal_article for more information
  - name: Matthias Gondan
    affiliation: University of Innsbruck
    address:
    - Department of Psychology
    - Innsbruck, Austria
    url: https://www.uibk.ac.at/psychologie/mitarbeiter/gondan-rochon/index.html.en
    orcid: 0000-0001-9974-0057
    email:  Matthias.Gondan-Rochon@uibk.ac.at
  - name: Irene Alfarone
    url: https://www.uibk.ac.at/psychologie/mitarbeiter/alfarone/index.html.en
    email: Irene.Alfarone@uibk.ac.at
    orcid: 0000-0002-8409-8900
    affiliation: University of Innsbruck
    address:
    - Department of Psychology
    - Innsbruck, Austria
type: package
output: 
  rjtools::rjournal_web_article:
    self_contained: yes
    toc: no
bibliography: RJreferences.bib
---

@@

library(mathml)
library(plotly)
library(ggplot2)

@@

The R extension of the markdown language [@Xie2020] enables reproducible
statistical reports with nice typesetting in HTML, Microsoft Word, and Latex.
Since recently [@R, version 4.2], R's manual pages include support for
mathematical expressions [@Sarkar2022], which already is a big improvement.
However, rules for the translation of R's built-in language elements to their
mathematical representation are still lacking. So far, R expressions such
as `pbinom(k, N, p)` are printed as they are; pretty mathematical formulae
like \(P_\mathrm{Bi}(X \le k; N, p)\) require explicit Latex commands, that
is, `P_\mathrm{Bi}\left(X \le k; N, p\right)`. Except for minimalistic examples,
these commands are tedious to type in and hard to read in their source code.

@@

A few more custom operators are shown in Table\ 1.

```{r, echo=FALSE}
op <- list(cross=quote(x %*% y), dot=quote(x %.% y), kronecker=quote(X %x% Y),
  'integer division'=quote(x %/% y))
m <- lapply(op, FUN=mathout, flags=list(cat=FALSE))

t <- cbind(Operator=as.character(op), Output=m)
knitr::kable(t, caption = "Custom operators in \\pkg{mathml}", 
             label="tab:custom")
```

@@

Schwarz' [-@schwarz1994] model refers to two stimuli A and B, presented either
alone or in combination (redundant stimuli), with the redundant stimuli being
presented either simultaneously or with onset asynchrony \(\tau\). The channel
activation is described as a two-dimensional Wiener process with
drifts \(\mu_\mathrm A\), \(\mu_\mathrm B\),
variances \(\sigma^2_\mathrm A\), \(\sigma^2_\mathrm B\),
correlation \(\rho_\mathrm{AB}\) and initial
conditions \(X_i(t=0)=0, i=\mathrm{A, B}\). The response is elicited when the
process reaches the absorbing barrier \(c > 0\) for the first time. In combined
stimuli ("redundant" stimuli), the overall diffusion process
is \(X_\mathrm{AB}(t) = X_\mathrm A(t) + X_\mathrm B(t)\) , which is again a
Wiener process with drift \(\mu_\mathrm A + \mu_\mathrm B\) and
variance \(\sigma^2_\mathrm A + \sigma^2_\mathrm B
  + 2\rho_\mathrm{AB}\sigma_\mathrm A\sigma_\mathrm B\).

In single-target trials, the first passages at \(c\) are expected at
\(E\left[D_i\right] = \frac{c}{\mu_i}, i = \mathrm{A, B}\), whereas in synchronous
redundant-target trials, we have
\(E[D_\mathrm{AB}]= \frac{c}{\mu_\mathrm A + \mu_\mathrm B}\).

For asynchronous stimuli, Schwarz [-@schwarz1994] derived the expected
first-passage time \(E[D_{\mathrm A(\tau)\mathrm B}]\) as a function of the
stimulus onset asyncrony \(\tau > 0\) (Eq.\ 10),

@@

The observable response time is assumed to be the sum of \(D\), the time
employed to reach the threshold for the decision, and \(M\), denoting the of the
other processes.

\(E[T]= E[D + M] = E[D] + E[M] = E[D] + \mu_M\)

@@

Schwarz [-@schwarz1994] applied the model to Miller’s [-@miller1986] data from
a redundant signals task with 13 onset
asynchronies \(0, \pm33, \pm67, \pm100, \pm133, \pm167, \pm\infty\) ms,
where 0 refers to synchronous AB, \(\infty\) refers to the single-target
presentation, and negative \(\tau\) denote those conditions in which B is
presented before A. Each condition was replicated 400 times. The 13 mean RTs and
standard deviations are given in Table 2 below.

Then, from Schwarz' [-@schwarz1994] Equations [2] [3] and [10], we calculated the expected first time passage for the respective \(\tau\) and the parameters \(\mu_ \mathrm A = 0.53, \mu_ \mathrm B = 1.32, \sigma_ \mathrm A = 4.43, \sigma_ \mathrm B = 18.3, c = 100, \mu_\mathrm M = 161\)

@@

For instance, we can see that with \(\tau = -167\) the expected first passage time is:

@@

To which we have to add \(\mu_M = 161\) (which represents the amount of time that the movement of providing the answer takes) and we obtain:

@@

Similarly, we obtain the first-time passage for every \(\tau\):

@@

Then we proceeded with the model fitting, using the data provided by Miller [-@miller1986] study in Schwarz [-@schwarz1994]. First, we calculated the predicted means RT as \(E[RT]=E[T]+\mu_M\).

We briefly remind that in diffusion models there is across-trial variability, and the drift rates are assumed to be normally distributed [@ratcliff2008] with at the first passage of \(c\) \(X(t) \sim N(\mu t , \sigma^2 t)\). Where \(T\) is expressed as \(T= \arg \min_t X(t)\geq c\) with \(T \sim IG (c, \mu, \sigma^2)\) with mean \(E(T)=\displaystyle \frac{c}{\mu}\) and variance \(Var(T)=\displaystyle \frac{c\sigma^2}{\mu^3}\). Furthermore, for the Central Limit Theorem \(\bar{T} = \displaystyle \frac{\Sigma_iT_i}{N} \dot\sim N (\displaystyle \frac{c}{\mu} , \displaystyle \frac{c\sigma^2}{\mu^3})\)

@@

\(X^2=\Sigma_\tau z^2_\tau \sim\chi^2(13-5\  dgf)\).

@@

The calculated Goodness of Fit value is \(X^2= 30.54\), \(p= 0.0002\). Additionally, we proceeded with the parameters' estimation with the Least Squares Method.

@@

: Table 1: Estimated Model's Parameters

| Parameters | Estimate | 95% Confidence Interval
| :---: | :---: | :---:|
| \( \mu_ \mathrm A \)| 0.53 | 0.51, 0.56
| \( \sigma_ \mathrm A \)| 4.4 | 3.8, 5.1
| \( \mu_ \mathrm B \)| 1.32 | 1.2, 1.4
| \( \sigma_ \mathrm B \)| 18.3 | 11.6, 24.9
| \( \mu_ \mathrm M \)| 161 | 157, 165

Furthermore, we can observe from the values reported in the Table below that, with few exceptions (e.i. \(\tau= 100\)) Miller's [-@miller1986] observed values are quite similar to the one predicted by Schwarz' [-@schwarz1994] model. Showing that, despite the calculated Goodness of Fit measure, the model proposed by Schwarz [-@schwarz1994] sufficiently fit the data.

: Table 2: Predicted Means RT and Observed Means RT and SDs from Miller's [-@miller1986] study in Schwarz [-@schwarz1994]

| SOA      | Predicted Mean RT | Observed Mean RT | Observed SD |
|:--------:|:-----------------:|:----------------:|:-----------:|
\(-\infty\)| 237 | 231 | 56 |
\(-167\)   | 231 | 234 | 58 |
\(-133\)   | 230 | 230 | 40 |
\(-100\)   | 228 | 227 | 40 |
\(-67\)    | 226 | 228 | 32 |
\(-33\)    | 222 | 221 | 28 |
\(0\)      | 215 | 217 | 28 |
\(33\)     | 239 | 238 | 28 |
\(67\)     | 262 | 263 | 26 |
\(100\)    | 283 | 277 | 30 |
\(133\)    | 300 | 298 | 32 |
\(167\)    | 313 | 316 | 34 |
\(\infty\) | 350 | 348 | 92 |

 _Note_.  Parameters: \(\mu_ \mathrm A = 0.53, \mu_ \mathrm B = 1.32, \sigma_ \mathrm A = 4.4, \sigma_ \mathrm B = 18.3, c = 100, \mu_\mathrm M = 161\)

>>>>

---
title: "mathml: Translate R expressions to MathML and LaTeX/MathJax"
date: "2022-12-09"
abstract: >
  This R\ package translates R\ objects to suitable entities in MathML or LaTeX,
  thereby enabling pretty mathematical representation of R\ functions or
  algorithms in data analysis, scientific reports and interactive web content.
  R\ code text descriptions and mathematical content already coexist,
  side-by-side, in the RMarkdown document rendering language. The present
  package allows to use the same R\ objects for both data analysis and
  typesetting in documents or web content. By tightening the link between the
  statistical analysis and its verbal or symbolic representation, this new
  feature is a futher step towards reproducible science. User-defined hooks
  allow to extend the package by mapping new variable names and functions to
  new MathML and LaTeX entities. We provide a few working examples that
  illustrate the use of the package in a little scientific report and in an
  interactive web page.
draft: true
author:  
  # see ?rjournal_article for more information
  - name: Matthias Gondan
    affiliation: University of Innsbruck
    address:
    - Department of Psychology
    - Innsbruck, Austria
    url: https://www.uibk.ac.at/psychologie/mitarbeiter/gondan-rochon/index.html.en
    orcid: 0000-0001-9974-0057
    email:  Matthias.Gondan-Rochon@uibk.ac.at
  - name: Irene Alfarone
    url: https://www.uibk.ac.at/psychologie/mitarbeiter/alfarone/index.html.en
    email: Irene.Alfarone@uibk.ac.at
    orcid: 0000-0002-8409-8900
    affiliation: University of Innsbruck
    address:
    - Department of Psychology
    - Innsbruck, Austria
type: package
output: 
  rjtools::rjournal_web_article:
    self_contained: yes
    toc: no
    tables: yes
bibliography: RJreferences.bib
---

@@

library(mathml)
library(plotly)
library(ggplot2)
library(kableExtra)

@@

The R extension of the markdown language [@Xie2020] enables reproducible
statistical reports with nice typesetting in HTML, Microsoft Word, and Latex.
Since recently [@R, version 4.2], R's manual pages include support for
mathematical expressions [@Sarkar2022], which already is a big improvement.
However, rules for the translation of R's built-in language elements to their
mathematical representation are still lacking. So far, R expressions such
as `pbinom(k, N, p)` are printed as they are; pretty mathematical formulae
like $P_\mathrm{Bi}(X \le k; N, p)$ require explicit Latex commands, that
is, `P_\mathrm{Bi}\left(X \le k; N, p\right)`. Except for minimalistic examples,
these commands are tedious to type in and hard to read in their source code.

@@

A few more custom operators are shown in Table \@ref(tab:custom-operators).

```{r custom-operators, echo=FALSE}
op <- list(cross=quote(x %*% y), dot=quote(x %.% y), kronecker=quote(X %x% Y),
  integerdivision=quote(x %/% y))
m <- lapply(op, FUN=mathjax, flags=list(cat=FALSE))
t <- cbind(Operator=as.character(op), Output=m)
knitr::kable(t, caption = "Custom operators in mathml")
```

@@

Schwarz' [-@schwarz1994] model refers to two stimuli A and B, presented either
alone or in combination (redundant stimuli), with the redundant stimuli being
presented either simultaneously or with onset asynchrony $\tau$. The channel
activation is described as a two-dimensional Wiener process with
drifts $\mu_\mathrm A$, $\mu_\mathrm B$,
variances $\sigma^2_\mathrm A$, $\sigma^2_\mathrm B$,
correlation $\rho_\mathrm{AB}$ and initial
conditions $X_i(t=0)=0, i=\mathrm{A, B}$. The response is elicited when the
process reaches the absorbing barrier $c > 0$ for the first time. In combined
stimuli ("redundant" stimuli), the overall diffusion process
is $X_\mathrm{AB}(t) = X_\mathrm A(t) + X_\mathrm B(t)$ , which is again a
Wiener process with drift $\mu_\mathrm A + \mu_\mathrm B$ and
variance $\sigma^2_\mathrm A + \sigma^2_\mathrm B
  + 2\rho_\mathrm{AB}\sigma_\mathrm A\sigma_\mathrm B$.

In single-target trials, the first passages at $c$ are expected at
$E\left[D_i\right] = \frac{c}{\mu_i}, i = \mathrm{A, B}$, whereas in synchronous
redundant-target trials, we have
$E[D_\mathrm{AB}]= \frac{c}{\mu_\mathrm A + \mu_\mathrm B}$.

For asynchronous stimuli, Schwarz [-@schwarz1994] derived the expected
first-passage time $E[D_{\mathrm A(\tau)\mathrm B}]$ as a function of the
stimulus onset asyncrony $\tau > 0$ (Eq.\ 10),

@@

The observable response time is assumed to be the sum of $D$, the time
employed to reach the threshold for the decision, and $M$, denoting the of the
other processes.

$E[T]= E[D + M] = E[D] + E[M] = E[D] + \mu_M$

@@

Schwarz [-@schwarz1994] applied the model to Miller’s [-@miller1986] data from
a redundant signals task with 13 onset
asynchronies $0, \pm33, \pm67, \pm100, \pm133, \pm167, \pm\infty$ ms,
where 0 refers to synchronous AB, $\infty$ refers to the single-target
presentation, and negative $\tau$ denote those conditions in which B is
presented before A. Each condition was replicated 400 times. The 13 mean RTs and
standard deviations are given in Table 2 below.

Then, from Schwarz' [-@schwarz1994] Equations [2] [3] and [10], we calculated the expected first time passage for the respective $\tau$ and the parameters $\mu_ \mathrm A = 0.53, \mu_ \mathrm B = 1.32, \sigma_ \mathrm A = 4.43, \sigma_ \mathrm B = 18.3, c = 100, \mu_\mathrm M = 161$

@@

For instance, we can see that with $\tau = -167$ the expected first passage time is:

@@

To which we have to add $\mu_M = 161$ (which represents the amount of time that the movement of providing the answer takes) and we obtain:

@@

Similarly, we obtain the first-time passage for every $\tau$:

@@

Then we proceeded with the model fitting, using the data provided by Miller [-@miller1986] study in Schwarz [-@schwarz1994]. First, we calculated the predicted means RT as $E[RT]=E[T]+\mu_M$.

We briefly remind that in diffusion models there is across-trial variability, and the drift rates are assumed to be normally distributed [@ratcliff2008] with at the first passage of $c$ $X(t) \sim N(\mu t , \sigma^2 t)$. Where $T$ is expressed as $T= \arg \min_t X(t)\geq c$ with $T \sim IG (c, \mu, \sigma^2)$ with mean $E(T)=\displaystyle \frac{c}{\mu}$ and variance $Var(T)=\displaystyle \frac{c\sigma^2}{\mu^3}$. Furthermore, for the Central Limit Theorem $\bar{T} = \displaystyle \frac{\Sigma_iT_i}{N} \dot\sim N (\displaystyle \frac{c}{\mu} , \displaystyle \frac{c\sigma^2}{\mu^3})$

@@

$X^2=\Sigma_\tau z^2_\tau \sim\chi^2(13-5\  dgf)$.

@@

The calculated Goodness of Fit value is $X^2= 30.54$, $p= 0.0002$. Additionally, we proceeded with the parameters' estimation with the Least Squares Method.

@@

```{r}
# Table 1: Estimated Model's Parameters
#
# | Parameters | Estimate | 95% Confidence Interval
# | :---: | :---: | :---:|
# | $ \mu_ \mathrm A $| 0.53 | 0.51, 0.56
# | $ \sigma_ \mathrm A $| 4.4 | 3.8, 5.1
# | $ \mu_ \mathrm B $| 1.32 | 1.2, 1.4
# | $ \sigma_ \mathrm B $| 18.3 | 11.6, 24.9
# | $ \mu_ \mathrm M $| 161 | 157, 165
```

Furthermore, we can observe from the values reported in the Table below that, with few exceptions (e.i. $\tau= 100$) Miller's [-@miller1986] observed values are quite similar to the one predicted by Schwarz' [-@schwarz1994] model. Showing that, despite the calculated Goodness of Fit measure, the model proposed by Schwarz [-@schwarz1994] sufficiently fit the data.

```{r}
# : Table 2: Predicted Means RT and Observed Means RT and SDs from Miller's [-@miller1986] study in Schwarz [-@schwarz1994]
#
# | SOA      | Predicted Mean RT | Observed Mean RT | Observed SD |
# |:--------:|:-----------------:|:----------------:|:-----------:|
# $-\infty$| 237 | 231 | 56 |
# $-167$   | 231 | 234 | 58 |
# $-133$   | 230 | 230 | 40 |
# $-100$   | 228 | 227 | 40 |
# $-67$    | 226 | 228 | 32 |
# $-33$    | 222 | 221 | 28 |
# $0$      | 215 | 217 | 28 |
# $33$     | 239 | 238 | 28 |
# $67$     | 262 | 263 | 26 |
# $100$    | 283 | 277 | 30 |
# $133$    | 300 | 298 | 32 |
# $167$    | 313 | 316 | 34 |
# $\infty$ | 350 | 348 | 92 |
#
# _Note_.  Parameters: $\mu_ \mathrm A = 0.53, \mu_ \mathrm B = 1.32, \sigma_ \mathrm A = 4.4, \sigma_ \mathrm B = 18.3, c = 100, \mu_\mathrm M = 161$
```
